The Astropy Project

### 2017-11-14 main.tex
change description to subsubsection to avoid awkward layout

The \astropypkg has a regular release schedule consisting of new significant
releases every 6 months, with bugfix releases as needed \citep{ape2}.
The major releases will contain new features or any significant changes, whereas
the bugfix releases only contain fixes to code or documentation but no new
features.
Some versions are additionally designated as ``Long-term support'' (LTS)
releases, which continue to receive bugfixes for 2 years following the release
with no changes to the API.
The LTS versions are ideal for pipelines and other applications where API
stability is essential.
The latest LTS release (v2.0) is also the last one that supports \python 2, and
will receive bug fixes until the end of 2019 \citep{ape10}.

@@

The \astropypkg core package is available from several of the \python distributions for scientific
% * <mattwcraig@gmail.com> 2017-10-27T07:13:51.093Z:
%
% > \astropypkg i
% I changed a few of the macros in here to refer to the package rather than the project because it is the package being discussed here.
%
% ^.
computing, such as Anaconda or Enthought Canopy, which include also some of
\astropy's affiliated packages.
Alternatively, \astropypkg can be readily installed from the Python Package
Index\footnote{\url{https://pypi.python.org/pypi}} (PyPI) using pip, the
recommended command-line tool for installing \python packages
that is included by default since \python 2 $\geq$ 2.7.9 and \python 3 $\geq$ 3.4.
For most recent \astropypkg releases, pre-built wheel distributions are obtainable
from PyPI for Windows, Mac OS X and Linux systems that provide faster
installation compared to the source distribution.  Optionally, the version of
the package and the dependencies needed to run or build the software are given
by a specifier to the pip package manager as described in PEP 440\footnote{For
more detailed installation instructions and optional flags see
\url{http://docs.astropy.org/en/stable/install.html\#installing-astropy}}.

Packaged versions of \astropy are also available for several
operating systems that are managed by a system's package manager,
including Debian, Ubuntu, Fedora, Gentoo, Arch Linux, Mac OS X (e.g.,
via brew, macports or fink) and FreeBSD.  However, on occasion these packages
may be outdated compared with the latest stable release of \astropy accessible
from the PyPI. Pre-built binary conda packages for all of the affiliated packages are also available  via the \texttt{astropy} conda channel for the three major platforms.
% * <mattwcraig@gmail.com> 2017-10-27T07:16:27.310Z:
%
% > re-built binary conda packages
% Feel free to remove if this is too detailed. I do think the available of the whole ecosystem across all platforms has helped speed its adoption.
%
% ^.

@@

\inlinecomment{BMS}{The itemizations below look very weird being boldface under a much less prominent subsection title. would it be nicer in the final version if we leave it as is in the draft?}

@@

The \texttt{astropy.units} subpackage adds support for representing units and
numbers with associated units --- ``quantities'' --- in code.
Historically, quantities in code have often been represented simply as numbers,
with units implied or noted via comments in the code because of considerations
about speed: having units associated with numbers inherently adds overhead to
numerical operations.
In \texttt{astropy.units}, \texttt{Quantity} objects subclass the \texttt{numpy}
array object and has been designed with speed in mind.
As of \astropypkg version 2.0, units and quantities are prevalent in most other
\astropypkg subpackages and are thus a key concept for using the package as a
whole.
Units are intimately entwined in the definition of astronomical coordinates and
thus nearly all functionality in the \texttt{astropy.coordinates} subpackage
(see \sectionname~\ref{sec:coordinates}) depends on this functionality.
For most other subpackages, quantities are at least accepted, and often expected
by default.

@@

\begin{description}
    \item[Interaction with \package{numpy} arrays] The \texttt{Quantity} object
        is now a subclass of the \texttt{numpy.ndarray} object and therefore
        works well with many of the functions in \texttt{numpy} that support
        array operations. For example, \texttt{Quantity} objects with angular
        units can be passed in to the trigonometric functions implemented in
        \texttt{numpy} (e.g., \texttt{numpy.cos}) and do the right thing: the
        units are internally converted to radians (what the \texttt{numpy}
        trigonometric functions expect) before being passed to \texttt{numpy}.
    \item[Speed improvements] The \texttt{Quantity} object is now significantly
        faster than in early versions of \astropypkg. This is in large part due
        to the fact that \texttt{Quantity} subclasses the \texttt{numpy.ndarray}
        object, but even relative to more recent versions of \astropypkg,
        operations with \texttt{Quantity} objects are generally faster. This is
        largely thanks to improvements in the way \texttt{numpy} arrays can be
        subclassed.
    \item[Logarithmic units and magnitudes] By default, taking the logarithm of
        a \texttt{Quantity} object with non-dimensionless units intentionally
        fails.
        However, some well-known units are actually logarithmic quantities,
        where the logarithm of the value is taken with respect to some reference
        value.
        Examples include astronomical magnitudes, which are logarithmic fluxes,
        and decibels, which are more generic logarithmic ratios of quantities.
        Logarithmic, relative units are now supported in \texttt{astropy.units}.
    \item[Using quantities in tables] The \texttt{astropy.table} subpackage (see
        \sectionname~\ref{sec:table}) now supports using quantities in tables
        with the \texttt{QTable} class.
        As a result, units specified in FITS tables are supported, and tables
        with quantities can be read and written to various file formats.
    \item[Defining functions that require quantities] When writing code or
        functions that expect \texttt{Quantity} objects, we often want to
        enforce that the input units have the correct type.
        For example, we may want to require only length-type \texttt{Quantity}
        objects.
        These requirements often lead to implementing repetitive code for
        validating \texttt{Quantity} inputs.
        \texttt{astropy.units} now provides a \python decorator,
        \texttt{quantity\_input()}, that does this verification automatically.
\end{description}

@@

A key concept behind the design of this subpackage is that coordinate
\textit{representations} and \textit{reference systems / frames} are independent
of one another.
For example, a set of coordinates in the International Celestial Reference
System (ICRS) reference frame could be represented as spherical (right
ascension, declination, and distance from solar system barycenter) or Cartesian
coordinates ($x$, $y$, $z$ with the origin at ).
They can therefore change representations independent of being transformed to
other reference frames (e.g., the Galactic coordinate frame).

@@

\begin{description}
    \item[Local Earth coordinate frames] In addition to representing celestial
    coordinates, \astropypkg now supports specifying positions on the Earth in
    a number of different geocentric systems with the \texttt{EarthLocation}
    class.
    With this, \astropypkg now supports Earth-location-specific coordinate
    systems such as the altitude-azimuth (\texttt{AltAz}) or horizontal system.
    Transformations between \texttt{AltAz} and any Barycentric coordinate frame
    also requires specifying a time using the \texttt{Time} class from
    \texttt{astropy.time}.
    With this new functionality, many of the common tasks associated with
    observation planning can now be completed with \astropypkg or the
    \astropy-affiliated package \package{astroplan}\citep{astroplan_AAS}.

    \item[Proper motion and velocity transformations]
    In addition to positional coordinate data, the \texttt{Frame} classes now
    also support velocity data.
    As the default representation for most frames is spherical, most of the
    \texttt{Frame} classes expect proper motion and radial velocity components
    to specify the velocity information.
    The names of the proper motion components all start with \texttt{pm} and
    adopt the same longitude and latitude names as the positional components.
    Transforming coordinates with velocity data is also supported, but in some
    cases the transformed velocity components have limited accuracy because the
    transformations are done numerically.
    The visualization of the coordinate frame transform graph highlights which
    velocity transformations can be done exactly and which transformations are
    done using a finite-difference scheme.
    The low-level interface for specifying and transforming velocity data (see
    the next point) is currently experimental.
    As such, in v2.0, only the \texttt{Frame} classes (and not the
    \texttt{SkyCoord} class) support handling velocities.

    \item[Derivatives of coordinate representations]
    As mentioned above, the \texttt{Representation} classes act like
    three-dimensional vectors.
    \texttt{astropy.coordinates} now also supports handling first derivatives of
    vectors / representations through the new \texttt{Differential} classes.
    The \texttt{Differential} classes are currently used internally within the
    \texttt{Frame} classes to store the velocity data.

    \item[Solar System Ephemerides]
    Also new is support for computing ephemerides of major solar system bodies
    and outputting the resulting positions as coordinate objects.
    These ephemerides can be computed either using analytic approximations from
    ERFA, or from downloaded JPL ephemerides (the latter requires the
    \package{jplephem}\footnote{\url{https://github.com/brandon-rhodes/python-jplephem}}
    optional dependency and an internet connection).
    The \texttt{get\_body} and \texttt{get\_body\_barycentric} functions provide
    the gateway to this functionality, yielding \texttt{SkyCoord} objects given
    a specific body and choice of ephemeris source.

\end{description}

@@

\begin{description}
    \item[Barycentric and Heliocentric corrections] Detailed eclipse or transit
        timing requires accounting for light travel time differences from the
        source to the observatory because of the Earth's motion.
        It is therefore common to instead convert times to the Solar System
        barycenter or heliocenter where the relative timing of photons is
        standardized.
        With the location of a source on the sky (i.e. a \texttt{SkyCoord}
        object), the location of an observatory on Earth (i.e. an
        \texttt{EarthLocation} object), and time values as \texttt{Time}
        objects, the time corrections to shift to the solar system barycenter or
        heliocenter can now be computed with \package{astropy.time} using the
        \texttt{light\_travel\_time} method of a \texttt{Time} object.
    \item[\texttt{Time} objects as \texttt{Table} columns] \texttt{Time} objects
        can now be added as columns in \package{astropy.table} \texttt{Table}
        objects. This is described in more detail in
        \sectionname~\ref{sec:table} with the concept of a ``mixin'' column
        type.
\end{description}

@@

We summarize key new features or updates to \package{astropy.table} below:
\begin{description}
    \item[Support for grouped table operations] Some tables contain data that
        naturally form groups.
        For example, a table of photometry for galaxies, stars, and quasars.
        We may then want to split the table or a column into these groups and
        compute some statistic on the data, or filter the groups based on some
        criteria.
        Grouped operations are now supported by \texttt{Table} objects,
        mainly through the \texttt{group\_by()} method and subsequent operations
        on the grouped version of the table or column.
    \item[Support for table concatenation] Tables can be combined in several
        different ways.
        If two tables have the same rows, we may want to stack them
        ``vertically'' to create a new table with the same columns but all rows.
        If two tables are row-matched but have distinct columns, we may want to
        stack them ``horizontally'' to create a new table with the same rows
        but all columns.
        For other situations, more general table concatenations or joins are
        also possible when two tables share some common columns.
        All three of these operations are now possible with \texttt{Table}
        objects through the \texttt{vstack}, \texttt{hstack}, and \texttt{join}
        functions in \package{astropy.table}.
    \item[Mixin columns for time and coordinates] The default data containers
        for \texttt{Table} objects are \texttt{Column} objects.
        However, sometimes the data for a given column is better represented as
        a \texttt{Quantity}, or a sky coordinate (\texttt{SkyCoord}), or an
        astronomical \texttt{Time} object.
        The \texttt{Table} object now supports these and other objects to serve
        as array-like containers for column data through the ``mixin'' protocol.
    \item[Transform to and from \texttt{Pandas} \texttt{DataFrame} objects]
        \texttt{Table} objects can now be created from a \texttt{DataFrame}
        object (\texttt{from\_pandas()}), or exported to a \texttt{DataFrame}
        object (\texttt{to\_pandas()}).
        \texttt{DataFrame} objects do not support multi-dimensional array
        columns, so this conversion will fail if the source \texttt{Table}
        object contains any such columns.
\end{description}

@@

We summarize key new features or updates to \package{astropy.io} below,
organized by sub-subpackage:
\begin{description}

    \item [Enhanced CSV format]

@@

    \item[\texttt{.ascii}: fast readers and writers]

@@

    \item[\texttt{.ascii}: HTML tables]

@@

    \item [\texttt{.fits}: command-line scripts (\texttt{fitsheader}, \texttt{fitsinfo})]

@@

    \item [\texttt{.fits}: lazy loading ]

@@

    \item [\texttt{.misc}: YAML serialization ]

@@

\end{description}

>>>>

The \astropypkg package has a regular release schedule consisting of new significant
releases every 6 months, with bugfix releases as needed \citep{ape2}.
The major releases contain new features or any significant changes, whereas
the bugfix releases only contain fixes to code or documentation but no new
features.
Some versions are additionally designated as ``Long-term support'' (LTS)
releases, which continue to receive bugfixes for 2 years following the release
with no changes to the API.
The LTS versions are ideal for pipelines and other applications where API
stability is essential.
The latest LTS release (v2.0) is also the last one that supports \python 2, and
will receive bug fixes until the end of 2019 \citep{ape10}.

@@

The \astropypkg core package is available from several of the \python distributions for scientific
% * <mattwcraig@gmail.com> 2017-10-27T07:13:51.093Z:
%
% > \astropypkg i
% I changed a few of the macros in here to refer to the package rather than the project because it is the package being discussed here.
%
% ^.
computing, such as Anaconda or Enthought Canopy, which include also some of
\astropy's affiliated packages.
Alternatively, \astropypkg can be readily installed from the Python Package
Index\footnote{\url{https://pypi.python.org/pypi}} (PyPI) using pip, the
recommended command-line tool for installing \python packages
that is included by default since \python 2 $\geq$ 2.7.9 and \python 3 $\geq$ 3.4.
For most recent \astropypkg releases, pre-built wheel distributions are obtainable
from PyPI for Windows, Mac OS X and Linux systems that provide faster
installation compared to the source distribution.
% hamogu: I think that's too detailed for a journal paper.
%     It's in the docs if anyone needs it.
%Optionally, the version of
%the package and the dependencies needed to run or build the software are given
%by a specifier to the pip package manager as described in PEP 440\footnote{For
%more detailed installation instructions and optional flags see
%\url{http://docs.astropy.org/en/stable/install.html\#installing-astropy}}.

Packaged versions of \astropy are also available for several
operating systems that are managed by a system's package manager,
including Debian, Ubuntu, Fedora, Gentoo, Arch Linux, Mac OS X (e.g.,
via brew, macports or fink) and FreeBSD.  However, on occasion these packages
may be outdated compared with the latest stable release of \astropy accessible
from the PyPI. Pre-built binary conda packages for all of the affiliated packages are also available  via the \texttt{astropy} conda channel for Linux, Max OS X, and MS Windows.
% * <mattwcraig@gmail.com> 2017-10-27T07:16:27.310Z:
%
% > re-built binary conda packages
% Feel free to remove if this is too detailed. I do think the available of the whole ecosystem across all platforms has helped speed its adoption.
%
% ^.

@@

@@

The \texttt{astropy.units} subpackage adds support for representing units and
numbers with associated units --- ``quantities'' --- in code.
Historically, quantities in code have often been represented simply as numbers,
with units implied or noted via comments in the code because of considerations
about speed: having units associated with numbers inherently adds overhead to
numerical operations.
In \texttt{astropy.units}, \texttt{Quantity} objects subclass the \texttt{numpy}
array object and have been designed with speed in mind.
As of \astropypkg version 2.0, units and quantities are prevalent in most other
\astropypkg subpackages and are thus a key concept for using the package as a
whole.
Units are intimately entwined in the definition of astronomical coordinates and
thus nearly all functionality in the \texttt{astropy.coordinates} subpackage
(see \sectionname~\ref{sec:coordinates}) depends on this functionality.
For most other subpackages, quantities are at least accepted, and often expected
by default.

@@

\subsubsection{Interaction with \package{numpy} arrays}

The \texttt{Quantity} object
        is now a subclass of the \texttt{numpy.ndarray} object and therefore
        works well with many of the functions in \texttt{numpy} that support
        array operations. For example, \texttt{Quantity} objects with angular
        units can be passed in to the trigonometric functions implemented in
        \texttt{numpy} (e.g., \texttt{numpy.cos}) and do the right thing: the
        units are internally converted to radians (what the \texttt{numpy}
        trigonometric functions expect) before being passed to \texttt{numpy}.

\subsubsection{Speed improvements} The \texttt{Quantity} object is now significantly
        faster than in early versions of \astropypkg. This is in large part due
        to the fact that \texttt{Quantity} subclasses the \texttt{numpy.ndarray}
        object, but even relative to more recent versions of \astropypkg,
        operations with \texttt{Quantity} objects are generally faster. This is
        largely thanks to improvements in the way \texttt{numpy} arrays can be
        subclassed.

\subsubsection{Logarithmic units and magnitudes}
        By default, taking the logarithm of
        a \texttt{Quantity} object with non-dimensionless units intentionally
        fails.
        However, some well-known units are actually logarithmic quantities,
        where the logarithm of the value is taken with respect to some reference
        value.
        Examples include astronomical magnitudes, which are logarithmic fluxes,
        and decibels, which are more generic logarithmic ratios of quantities.
        Logarithmic, relative units are now supported in \texttt{astropy.units}.

\subsubsection{Using quantities in tables}

        The \texttt{astropy.table} subpackage (see
        \sectionname~\ref{sec:table}) now supports using quantities in tables
        with the \texttt{QTable} class.
        As a result, units specified in FITS tables are supported, and tables
        with quantities can be read and written to various file formats.

\subsubsection{Defining functions that require quantities}
        When writing code or
        functions that expect \texttt{Quantity} objects, we often want to
        enforce that the input units have the correct type.
        For example, we may want to require only length-type \texttt{Quantity}
        objects.
        These requirements often lead to implementing repetitive code for
        validating \texttt{Quantity} inputs.
        \texttt{astropy.units} now provides a \python decorator,
        \texttt{quantity\_input()}, that does this verification automatically.

@@

A key concept behind the design of this subpackage is that coordinate
\textit{representations} and \textit{reference systems / frames} are independent
of one another.
For example, a set of coordinates in the International Celestial Reference
System (ICRS) reference frame could be represented as spherical (right
ascension, declination, and distance from solar system barycenter) or Cartesian
coordinates ($x$, $y$, $z$ with the origin at barycenter).
They can therefore change representations independent of being transformed to
other reference frames (e.g., the Galactic coordinate frame).

@@

\subsubsection{Local Earth coordinate frames}
In addition to representing celestial
    coordinates, \astropypkg now supports specifying positions on the Earth in
    a number of different geocentric systems with the \texttt{EarthLocation}
    class.
    With this, \astropypkg now supports Earth-location-specific coordinate
    systems such as the altitude-azimuth (\texttt{AltAz}) or horizontal system.
    Transformations between \texttt{AltAz} and any Barycentric coordinate frame
    also requires specifying a time using the \texttt{Time} class from
    \texttt{astropy.time}.
    With this new functionality, many of the common tasks associated with
    observation planning can now be completed with \astropypkg or the
    \astropy-affiliated package \package{astroplan}\citep{astroplan_AAS}.

\subsubsection{Proper motion and velocity transformations}
    In addition to positional coordinate data, the \texttt{Frame} classes now
    also support velocity data.
    As the default representation for most frames is spherical, most of the
    \texttt{Frame} classes expect proper motion and radial velocity components
    to specify the velocity information.
    The names of the proper motion components all start with \texttt{pm} and
    adopt the same longitude and latitude names as the positional components.
    Transforming coordinates with velocity data is also supported, but in some
    cases the transformed velocity components have limited accuracy because the
    transformations are done numerically.
    The visualization of the coordinate frame transform graph highlights which
    velocity transformations can be done exactly and which transformations are
    done using a finite-difference scheme.
    The low-level interface for specifying and transforming velocity data (see
    the next point) is currently experimental.
    As such, in v2.0, only the \texttt{Frame} classes (and not the
    \texttt{SkyCoord} class) support handling velocities.

\subsubsection{Derivatives of coordinate representations}
    As mentioned above, the \texttt{Representation} classes act like
    three-dimensional vectors.
    \texttt{astropy.coordinates} now also supports handling first derivatives of
    vectors / representations through the new \texttt{Differential} classes.
    The \texttt{Differential} classes are currently used internally within the
    \texttt{Frame} classes to store the velocity data.

\subsubsection{Solar System Ephemerides}
    Also new is support for computing ephemerides of major solar system bodies
    and outputting the resulting positions as coordinate objects.
    These ephemerides can be computed either using analytic approximations from
    ERFA, or from downloaded JPL ephemerides (the latter requires the
    \package{jplephem}\footnote{\url{https://github.com/brandon-rhodes/python-jplephem}}
    optional dependency and an internet connection).
    % hamogu: I think the specific names of functions can be in the docs
    % and don't need to be repeated here.
    %The \texttt{get\_body} and \texttt{get\_body\_barycentric} functions provide
    %the gateway to this functionality, yielding \texttt{SkyCoord} objects given
    %a specific body and choice of ephemeris source.

@@

\subsubsection{Barycentric and Heliocentric corrections}
Detailed eclipse or transit
        timing requires accounting for light travel time differences from the
        source to the observatory because of the Earth's motion.
        It is therefore common to instead convert times to the Solar System
        barycenter or heliocenter where the relative timing of photons is
        standardized.
        With the location of a source on the sky (i.e. a \texttt{SkyCoord}
        object), the location of an observatory on Earth (i.e. an
        \texttt{EarthLocation} object), and time values as \texttt{Time}
        objects, the time corrections to shift to the solar system barycenter or
        heliocenter can now be computed with \package{astropy.time} using the
        \texttt{light\_travel\_time} method of a \texttt{Time} object.
\subsubsection{\texttt{Time} objects as \texttt{Table} columns} \texttt{Time} objects
        can now be added as columns in \package{astropy.table} \texttt{Table}
        objects. This is described in more detail in
        \sectionname~\ref{sec:table} with the concept of a ``mixin'' column
        type.

@@

Next, we summarize key new features or updates to \package{astropy.table}.

\subsubsection{Support for grouped table operations}
Tables can contain data that naturally form groups, for example monitoring of an object in different bands.
We may then want to split the table or a column into these groups (in this example: observations in different filters)  and
compute some statistic on the data, or filter the groups based on some
criteria.
Grouped operations are now supported by \texttt{Table} objects.
%,
%mainly through the \texttt{group\_by()} method and subsequent operations
%on the grouped version of the table or column.

\subsubsection{Support for table concatenation}
Tables can be combined in several
different ways.
If two tables have the same rows, we may want to stack them
``vertically'' to create a new table with the same columns but all rows.
If two tables are row-matched but have distinct columns, we may want to
stack them ``horizontally'' to create a new table with the same rows
but all columns.
For other situations, more general table concatenations or joins are
also possible when two tables share some common columns.
%All three of these operations are now possible with \texttt{Table}
%objects through the \texttt{vstack}, \texttt{hstack}, and \texttt{join}
%functions in \package{astropy.table}.

\subsubsection{Mixin columns for time and coordinates}
The \texttt{Table} object now supports \texttt{Quantity}, sky coordinate (\texttt{SkyCoord}),
astronomical \texttt{Time} objects and and other objects to serve
as array-like containers for column data through the ``mixin'' protocol.

%hamogu: I've commented this out.
% I think it's very specific and not relevant to non-developers.
% If we want to keep it, we need to add at least 102 senteces about what Pandas
% is and I think that's more text than is warrented for this feature.

%\subsubsection{Transform to and from \texttt{Pandas} \texttt{DataFrame} objects}
%\texttt{Table} objects can now be created from a \texttt{DataFrame}
%object (\texttt{from\_pandas()}), or exported to a \texttt{DataFrame}
%object (\texttt{to\_pandas()}).
%\texttt{DataFrame} objects do not support multi-dimensional array
%columns, so this conversion will fail if the source \texttt{Table}
%object contains any such columns.

@@

We summarize key new features or updates to \package{astropy.io} below,
organized by sub-subpackage.

\subsubsection{Enhanced CSV format}

@@

\subsubsection{\texttt{.ascii}: fast readers and writers}

@@

\subsubsection{\texttt{.ascii}: HTML tables}

@@

\subsubsection{\texttt{.fits}: command-line scripts (\texttt{fitsheader}, \texttt{fitsinfo})}

@@

\subsubsection{\texttt{.fits}: lazy loading }

@@

\subsubsection{\texttt{.misc}: YAML serialization}

@@
